Benefits and risks associated with children’s and adolescents’ interactions with electronic screens: An umbrella review

### 2022-01-07 reports/manuscript_content.Rmd
Change to Papaja

---
title: "Benefits and risks associated with children’s and adolescents’ interactions with electronic screens: An umbrella review"
author:
- name: Taren Sanders
  affiliation: Institute for Positive Psychology and Education, ACU
  affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
  orcid_id: 0000-0002-4504-6008
- name: Michael Noetel
  affiliation: School of Health and Behavioural Sciences, ACU
  affiliation_url: https://www.acu.edu.au/about-acu/faculties-directorates-and-staff/faculty-of-health-sciences/school-of-behavioural-and-health-sciences
  orcid_id: 0000-0002-6563-8203
- name: Philip Parker
  affiliation: Institute for Positive Psychology and Education, ACU
  affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
  orcid_id: 0000-0002-4604-8566
- name: Borja Del Pozo Cruz
- name: Stuart Biddle
- name: Rimante Ronto
- name: Ryan Hulteen
- name: Rhiannon Parker
- name: George Thomas
- name: Katrien De Cocker
- name: Jo Salmon
- name: Kylie Hesketh
- name: Nicole Weeks
- name: Hugh Arnott
- name: Emma Devine
- name: Roberta Vasconcellos
- name: Chris Lonsdale
  affiliation: Institute for Positive Psychology and Education, ACU
  affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
  orcid_id: 000-0002-2523-5565
date: '`r format(Sys.Date())`'
repository_url: https://github.com/Motivation-and-Behaviour/screen_umbrella
bibliography: combined.bib
csl: vancouver-brackets.csl
params:
  render_img: true
  inc_ref: false
---

```{r setup, include=FALSE}
# This allows distill to render from a targets pipeline
Sys.setenv("RSTUDIO_VERSION" = '1.4.1725')

knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

# Uncomment to run outside of pipeline
# library(tidyverse)
# library(targets)
# library(xfun)
```

@@

Two researchers independently extracted all quantitative meta-analytic effect sizes, including moderation results. Where possible, they also extracted effect sizes from primary studies included in each meta-analysis. To facilitate comparisons, we converted effect sizes to Pearson’s $r$ using established formulae [@bonettTransformingOddsRatios2007;@bowmanEffectSizesStatistical2012;@jacobsEstimationBiserialCorrelation2017]. We excluded relative risk ratios from this conversion because meta-analyses did not contain sufficient information to do so. Effect sizes on the original metric are provided in supplementary materials.

@@

We initially planned to include systematic reviews without meta-analyses in a narrative summary alongside the main meta-analytic findings. However, we determined that combining results from the meta-analyses provided a robust overview. Readers interested in the relevant systematic reviews (i.e., without meta-analysis) can consult the list of references in eSupplementary File X.

@@

The most frequently reported exposures were `r knitr::combine_words(freq_exposures$out[1:4])`. Supplementary file X provides a list of all exposures identified. The most frequently reported outcomes were `r knitr::combine_words(freq_outcomes$out[1:5])`. In most cases (`r paste0(sum(freq_combos$n==1), "/", nrow(effects_clean))`), there was only one exposure/outcome combination, with `r sum(freq_combos$n==2)` appearing twice, and `r sum(freq_combos$n>2)` appearing three or more times. Full characteristics of the included studies are provided in Table \@ref(tab:desctable). Our process yielded `r unique_effects` unique effects/outcome combinations contributed from `r prisma$data$value[9]` reviews. These effects represent the findings of `r format(effects_sum$sum_k, big.mark=",")` primary studies comprised of `r format(effects_sum$sum_n, big.mark=",")` participants.

```{r desctable, eval=params$render_img, layout="l-screen-inset shaded"}
reviews_tables[[1]]$table
```

@@

reviews_count <- qual_data %>%
  mutate(
    across(
      eligibility_criteria_predefined_and_specified:heterogeneity_assessed,
      as.factor
    )
  ) %>%
  rowwise() %>%
  mutate(count_low = sum(
    c_across(
      eligibility_criteria_predefined_and_specified:heterogeneity_assessed
    ) == "low"
  ),
  count_high = sum(
    c_across(
      eligibility_criteria_predefined_and_specified:heterogeneity_assessed
    ) == "high"
  ))

@@

@@

@@

dualqualstr <- paste0(
  qual_str(n_high_dualquality, nrow(qual_data), " high risk ", suffix= "; "),
  qual_str(n_unclear_dualquality, nrow(qual_data), " unclear  ", prefix = "n", suffix = ")"))

The quality of the included meta-analyses was mixed (see Table \@ref(tab:desctable)). Most assessed heterogeneity `r qual_str(n_low_hetero, nrow(qual_data), " low risk ", suffix=" of meta-analyses)")`, reported the characteristics of the included studies `r qual_str(n_low_char, nrow(qual_data), " low risk ")`, and used a comprehensive and systematic search strategy `r qual_str(n_low_search, nrow(qual_data), " low risk ")`. Most reviews did not clearly report if their eligibility criteria was predefined `r qual_str(n_unclear_elig, nrow(qual_data), " unclear ")`. Many papers also did not complete dual independent screening of abstracts and full text `r qual_str(n_high_dualscreen, nrow(qual_data), " high risk ")` or did not clearly report the method of screening `r qual_str(n_unclear_dualscreen, nrow(qual_data), " unclear ")`. A similar trend was observed for dual independent quality assessment `r dualqualstr`. Overall, only `r all_low` meta-analyses were graded as low risk of bias on all criteria. `r all_high_txt` showed high risk of bias on all criteria.

@@

There were `r n_edu_effect` unique effects associated with education outcomes, including general learning outcomes, literacy, numeracy, and science. We removed `r n_edu_noindiv` effects that did not provide individual study-level data, `r n_edu_samplesize` effects with samples < 1,000, and `r n_edu_faileggers` with a significant Egger's test or insufficient studies to conduct the test. No remaining studies showed evidence of excessive significance. Effects not meeting one or more of these standards are presented in the supplementary material. The remaining `r n_edu_certain` met our criteria for statistical credibility and are described in Figure \@ref(fig:eduplot). These `r n_edu_certain` effects came from `r n_edu_unique` meta-analytic reviews analysing data from `r edu_summary$studies_n` empirical studies with `r format(edu_summary$total_n, big.mark=",")` individual participants.

@@

We identified `r n_health_effect` unique outcome-exposure combinations associated with health or health-related behaviour outcomes. We removed `r n_health_noindiv` effects that did not provide individual study-level data, `r n_health_samplesize` effects with samples < 1,000, and `r n_health_faileggers` effects with a significant Egger's test or insufficient studies to conduct the test. No remaining studies showed evidence of excessive significance. Effects not meeting one or more of these standards are presented in the supplementary material. The remaining `r n_health_certain` meta-analytic associations met our criteria for credible evidence and are described below (see also Figure \@ref(fig:healthplot)). These `r n_health_effect` effects came from `r n_health_unique` meta-analytic reviews analysing data from `r health_summary$studies_n` empirical studies with `r format(health_summary$total_n, big.mark=",")` individual participants.

@@

Digital advertising of unhealthy foods—both traditional advertising and video games developed by a brand for promotion—were associated with higher unhealthy food intake. Risky behaviors (e.g., sexual activity, risk taking, and substance abuse) were positively associated with social media use and sexy media use. General screen use was positively associated with depression. Television viewing was negatively correlated with sleep duration, but only at the 95% confidence level (weak evidence). All forms of screen use (general, television, and video games) were positively associated with body composition (e.g., BMI). Screen-based interventions which target health behaviours appeared effective.

@@

Across the health outcomes, most (`r n_health_p999` of `r n_health_certain` effects) were statistically significant at the 99.9\% level, with the remaining `r xfun::numbers_to_words(n_health_p95)` significant at 95\% confidence. However, all of the credible effects exhibited high levels of heterogeneity, with the lowest $I^2 = 75\%$. Additionally, most effects were small, with the association between unhealthy food advertising and intake the largest at $r = 0.25$ ($k = 13, N = 1,756$). Most of the effect sizes (`r n_health_r2`/`r n_health_certain`) had an absolute value of $r < 0.2$.

@@

**Note:** *Below are rough notes on ideas for the discussion. Please feel free to add ideas. We will revise once we agree on the rest of the paper.*

#### Broad summary of the findings relating back to the purpose
When meta-analysis examined general screen use, and did not specify the content, context or device, there was strong evidence showing potentially harmful associations with general learning, literacy, body composition, and depression. However, when meta-analyses included a more nuanced examination of exposures a more complex picture appeared. Many exposures were likely neither universally harmful nor consistently beneficial. For example, watching television programs was associated with poorer academic performance and learning. However, there was also evidence that if the content of the program was educational or the child watched the program with a parent (i.e., coviewing) then television viewing was associated with better literacy. Thus, parents play an important role in selecting content that is likely to benefit their children and, perhaps, interacting with their children in ways that may foster literacy (e.g., asking their children questions about the program). Similarly, the credible evidence we identified showed video game playing is associated with poorer body composition and learning. When the video game was designed specifically to teach numeracy, playing these games showed learning benefits. One might expect that video games designed to be physically active could confer health benefits, but none of the meta-analyses examining this hypothesis met our thresholds for credibility (see Supplementary File X).

@@

@@

# Appendix

#### Figures

* [Prisma Diagram](https://drive.google.com/file/d/1ezgeAr2WPhy6MNSESS8bhg3VMOrLAUC7/view?usp=sharing)
* [Education Outcomes](https://drive.google.com/file/d/1ODsKf3jjEGNCpUTz5tju8AHRpk_xIz-e/view?usp=sharing)
* [Health-Related Outcomes](https://drive.google.com/file/d/1wdKzliWLEFuVveAj6-MzOAsoqBhnkRJW/view?usp=sharing)
* [Supplemental Education Outcomes](https://drive.google.com/file/d/1wUotZRh-jvdBJh5dkstkMtvcz7DPn-vS/view?usp=sharing)
* [Supplemental Health-Related Outcomes](https://drive.google.com/file/d/1-3frrZ3woWjPqkO84Dn3_1CCaE8thIZn/view?usp=sharing)

#### Tables

* [Descriptive Table](https://drive.google.com/file/d/1t5XvwY2kzsOsbZ2FxbGPsYEecZz5PgD1/view?usp=sharing)
* Supplemental tables (TBC)

#### Supplementary Files

* [List of Exposures](https://drive.google.com/file/d/12rB0ihW9Hi1kxM-8AIBfjdCMO6TqP28n/view?usp=sharing)

#### Links

* [Manuscript Google Doc](https://docs.google.com/document/d/1NN4yJWQVepNjbX0mND4aidWWXcCDkg-eMepRnWdtQ9o/edit#)
* [Online Report](https://motivation-and-behaviour.github.io/screen_umbrella/)

# References

>>>>

@@

Two researchers independently extracted all quantitative meta-analytic effect sizes, including moderation results. Where possible, they also extracted effect sizes from primary studies included in each meta-analysis. To facilitate comparisons, we converted effect sizes to Pearson’s $r$ using established formulae [@bonettTransformingOddsRatios2007;@bowmanEffectSizesStatistical2012;@jacobsEstimationBiserialCorrelation2017]. We excluded relative risk ratios from this conversion because meta-analyses did not contain sufficient information to meaningfully convert. Effect sizes on the original metric are provided in supplementary materials.

@@

We initially planned to include systematic reviews without meta-analyses in a narrative summary alongside the main meta-analytic findings. However, we determined that combining results from the meta-analyses provided a robust overview. Readers interested in the relevant systematic reviews (i.e., without meta-analysis) can consult the list of references in [Supplementary File 1](https://drive.google.com/file/d/1JhMsySEyfby5MtcUcIVpp0LjDDfTA5IH/view?usp=sharing).

@@

The most frequently reported exposures were `r knitr::combine_words(freq_exposures$out[1:4])`. [Supplementary file 2](https://drive.google.com/file/d/12rB0ihW9Hi1kxM-8AIBfjdCMO6TqP28n/view?usp=sharing) provides a list of all exposures identified. The most frequently reported outcomes were `r knitr::combine_words(freq_outcomes$out[1:5])`. In most cases (`r paste0(sum(freq_combos$n==1), "/", nrow(effects_clean))`), there was only one exposure/outcome combination for an age group, with `r sum(freq_combos$n==2)` appearing twice, and `r sum(freq_combos$n>2)` appearing three or more times. Full characteristics of the included studies are provided in Table \@ref(tab:desctable). Our process yielded `r unique_effects` unique effects/outcome combinations contributed from `r prisma$data$value[9]` reviews. These effects represent the findings of `r format(effects_sum$sum_k, big.mark=",")` primary studies comprised of `r format(effects_sum$sum_n, big.mark=",")` participants.

```{r desctable, eval=params$inc_tabs, layout="l-screen-inset shaded"}
reviews_tables[[1]]$table
```

@@

reviews_count <- qual_data %>%
  mutate(
    across(
      eligibility_criteria_predefined_and_specified:heterogeneity_assessed,
      as.factor
    )
  ) %>%
  rowwise() %>%
  mutate(
    count_low = sum(
      c_across(
        eligibility_criteria_predefined_and_specified:heterogeneity_assessed
      ) == "low"
    ),
    count_high = sum(
      c_across(
        eligibility_criteria_predefined_and_specified:heterogeneity_assessed
      ) == "high"
    ),
    count_high_unclear = sum(
      c_across(
        eligibility_criteria_predefined_and_specified:heterogeneity_assessed
      ) %in% c("high", "unclear")
    )
  )

@@

all_high_unclear <- reviews_count %>% 
  filter(count_high_unclear==7) %>% 
  nrow()

@@

all_high_unclear_txt <- 
  paste0(numbers_to_words(all_high_unclear), 
         if (all_high_unclear>1 | all_high_unclear==0) " meta-analyses" else " meta-analysis")

@@

dualqualstr <- paste0(
  qual_str(n_high_dualquality, nrow(qual_data), " high risk ", suffix= "; "),
  qual_str(n_unclear_dualquality, nrow(qual_data), " unclear  ", prefix = "n", suffix = ")"))
```

The quality of the included meta-analyses was mixed (see Table \@ref(tab:desctable)). Most assessed heterogeneity `r qual_str(n_low_hetero, nrow(qual_data), " low risk ", suffix=" of meta-analyses)")`, reported the characteristics of the included studies `r qual_str(n_low_char, nrow(qual_data), " low risk ")`, and used a comprehensive and systematic search strategy `r qual_str(n_low_search, nrow(qual_data), " low risk ")`. Most reviews did not clearly report if their eligibility criteria was predefined `r qual_str(n_unclear_elig, nrow(qual_data), " unclear ")`. Many papers also did not complete dual independent screening of abstracts and full text `r qual_str(n_high_dualscreen, nrow(qual_data), " high risk ")` or did not clearly report the method of screening `r qual_str(n_unclear_dualscreen, nrow(qual_data), " unclear ")`. A similar trend was observed for dual independent quality assessment `r dualqualstr`. Overall, only `r all_low` meta-analyses were graded as low risk of bias on all criteria. `r all_high_txt` showed high risk of bias on all criteria (`r all_high_unclear_txt` if 'unclear' was included as 'high').

@@

There were `r n_edu_effect` unique effects associated with education outcomes, including general learning outcomes, literacy, numeracy, and science. We removed `r n_edu_noindiv` effects that did not provide individual study-level data, `r n_edu_samplesize` effects with samples < 1,000, and `r n_edu_faileggers` with a significant Egger's test or insufficient studies to conduct the test. No remaining studies showed evidence of excessive significance. Effects not meeting one or more of these standards are presented in [Supplementary File 3](https://drive.google.com/file/d/1wUotZRh-jvdBJh5dkstkMtvcz7DPn-vS/view?usp=sharing). The remaining `r n_edu_certain` met our criteria for statistical credibility and are described in Figure \@ref(fig:eduplot). These `r n_edu_certain` effects came from `r n_edu_unique` meta-analytic reviews analysing data from `r edu_summary$studies_n` empirical studies with `r format(edu_summary$total_n, big.mark=",")` individual participants.

@@

We identified `r n_health_effect` unique outcome-exposure combinations associated with health or health-related behaviour outcomes. We removed `r n_health_noindiv` effects that did not provide individual study-level data, `r n_health_samplesize` effects with samples < 1,000, and `r n_health_faileggers` effects with a significant Egger's test or insufficient studies to conduct the test. No remaining studies showed evidence of excessive significance. Effects not meeting one or more of these standards are presented in [Supplementary File 4](https://drive.google.com/file/d/1-3frrZ3woWjPqkO84Dn3_1CCaE8thIZn/view?usp=sharing) The remaining `r n_health_certain` meta-analytic associations met our criteria for credible evidence and are described below (see also Figure \@ref(fig:healthplot)). These `r n_health_effect` effects came from `r n_health_unique` meta-analytic reviews analysing data from `r health_summary$studies_n` empirical studies with `r format(health_summary$total_n, big.mark=",")` individual participants.

@@

Digital advertising of unhealthy foods—both traditional advertising and video games developed by a brand for promotion—were associated with higher unhealthy food intake. Risky behaviors (e.g., sexual activity, risk taking, and substance abuse) were positively associated with social media use and sexy media use. General screen use was positively associated with depression. Television viewing was negatively correlated with sleep duration, but only at the 95% confidence level (weak evidence). All forms of screen use (general, television, and video games) were positively associated with body composition (e.g., BMI), although the association was smaller for children than for adolescents or for combined populations. Screen-based interventions which target health behaviours appeared effective.

@@

Across the health outcomes, most (`r n_health_p999` of `r n_health_certain` effects) were statistically significant at the 99.9\% level, with the remaining `r xfun::numbers_to_words(n_health_p95)` significant at 95\% confidence. However, most of the credible effects exhibited high levels of heterogeneity, with all but one having $I^2 > 75\%$. Additionally, most effects were small, with the association between unhealthy food advertising and intake the largest at $r = 0.25$ ($k = 13, N = 1,756$). Most of the effect sizes (`r n_health_r2`/`r n_health_certain`) had an absolute value of $r < 0.2$.

@@

The primary goal of this review was to provide a holistic perspective on the influence of screens on children’s lives across a broad domain of outcomes. We found that when meta-analyses examined general screen use, and did not specify the content, context or device, there was strong evidence showing potentially harmful associations with general learning, literacy, body composition, and depression. However, when meta-analyses included a more nuanced examination of exposures a more complex picture appeared.

#### Broad summary of the findings relating back to the purpose

For example, watching television programs was associated with poorer academic performance and learning. However, there was also evidence that if the content of the program was educational or the child watched the program with a parent (i.e., coviewing) then television viewing was associated with better literacy. Thus, parents play an important role in selecting content that is likely to benefit their children and, perhaps, interacting with their children in ways that may foster literacy (e.g., asking their children questions about the program). Similarly, the credible evidence we identified showed video game playing is associated with poorer body composition and learning. When the video game was designed specifically to teach numeracy, playing these games showed learning benefits. One might expect that video games designed to be physically active could confer health benefits, but none of the meta-analyses examining this hypothesis met our thresholds for credibility (see Supplementary File X).

@@

Largely owing to small number of studies or missing individual study data, there were few age-based conclusions that could be drawn from reviews which met our criteria for statistical certainty. If we expand our view to include those reviews which did meet this threshold, there remains no clear trend. For example, advertising of unhealthy food was associated with unhealthy food choice for young children, but was not statistically significant for other age groups. Conversely, TV programs and movies were more strongly associated with lower physical activity for adolescents than for younger age groups.

@@
